---
title: "Regression Models: Motor Trend Cars Analysis"
author: "Mark Culp"
date: "April 10, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
This report examines Motor Trend car road tests extracted from the 1974 Motor Trend US magazine.  These road tests were conducted on 1973-1974 car models.  The tests examined 10 aspects of automobile design and performance for 32 different models.  We focused here on the miles per gallon (mpg) performance of automatic versus manual transmissions.   


# Appendix
```{r}

# Load data set
library(datasets)
data("mtcars")



```
## Exploratory Analysis

```{r}

# Examine columns and data types
str(mtcars)

# Range of MPG values
range(mtcars$mpg)

# Car transmissions: 
# -----------------
# 0 = automatic 
# 1 = manual 
table(mtcars$am)

```

## Fitting Multiple Models, Strategy

```{r}

```

## Residual Plot and Diagnostics
```{r}

```

## Conclusions : Quantified Uncertainty

Cars with manual transmissions have better gas mileage than cars with automatic transmissions.  This can be quantified by ...

```{r}

```


